### Where the data come from {.tabset .tabset-pills}

This report is based on a locally maintained dataset focused on WA State.  The data are updated about once each week, pulling, cleaning and merging data from two active national projects:  

* Mapping Police Violence (MPV, https://mappingpoliceviolence.us/) -- 2013+
* IncarcerNation (IN, https://incarcernation.com/) -- 2021+

We incorporate data from one legacy project:

* Fatal Encounters (https://fatalencounters.org/) -- 2000-2021

And the data from another legacy project are preserved in the cases from MPV:

* Washington Post (https://www.washingtonpost.com/graphics/investigations/police-shootings-database/) -- 2000-2025

In addition we conduct independent searches for incidents in WA state that these national projects may have missed, using a mix of google alerts and searches of WA-specific sources like medical examiner reports, independent investigation reports, social media, obituaries and public disclosure requests.

The primary data source is MPV, but due to coverage limitations in the MPV project (primarily the exclusion of most pursuit-related fatalities) we augment their data with cases from IncarcerNation, legacy data from Fatal Encounters and locally identified incidents.  Cases are linked and de-duplicated across source datasets, and all records in the final merged dataset include the original source case IDs.

____

> For the purposes of this report, we call the resulting dataset the WA local dataset.

____

#### Breakdown by source of data {.tabset .tabset-pills}

The breakdown of the data by source is shown in the tabs below. Many incidents are included in more than one dataset, some are included only in one, and there are multiple ways to present the coverage of each source and pattern of overlap.

##### Source coverage

The coverage of each data source can be assessed by the presence of the source ID tags by year, shown below.  The columns for each data source show the number of cases with that source ID tag.  The "Total Cases" column in the table represents the total cases in that year, not the sum across the data sources.  Comparing the number of cases in each of the Data Source columns to each other and the total gives a sense of what fraction of that year's cases were covered by each source, the overlap across the sources, and how that varies by year.

```{r tabTags}

tab <- homicides %>%
  group_by(Year = as.character(year)) %>%
  summarize(FE = sum(!is.na(feID)),
            MPV = sum(!is.na(mpvID)),
            WaPo = sum(!is.na(wapoID)),
            WaLocal = sum(!is.na(waID)),
            Total.Cases = n()
  ) %>%
  mutate(Percent = round(100*Total.Cases/sum(Total.Cases), 1)) %>%

  add_row(Year = "Total", summarise(., across(where(is.numeric), sum)))

tab %>%
  kable(caption = "Source Coverage by Year") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  add_header_above(header = c(" " = 1, "Data Source" = 4, "De-duplicated yearly totals"=2)) %>%
  row_spec(dim(tab)[1], bold = T)  %>%
  column_spec(which(names(tab)=="Cases"), bold=T) %>%
  column_spec(which(names(tab)=="Percent"), italic=T) %>%

  add_footnote(label = "Percents may not sum to 100 due to rounding",
               notation = "symbol")
```

##### Source overlap by year

Each case in the dataset can also be identified by the number of sources that include it, a direct measure of the overlapping sources of information.  The breakdown of source overlap by year is shown below. 

```{r tabUnique, message=F, warning=F}

tab <- homicides %>% 
  count(source.overlap, Year = as.character(year)) %>% 
  pivot_wider(names_from=source.overlap, values_from=n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.,0))) %>%
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  add_row(Year = "Total", summarise(., across(where(is.numeric), sum)))
  
tab %>%
  kableExtra::kbl(caption = "Breakdown of fatalities by number of data sources and year") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_header_above(header = c(" " = 1, "Data Source(s) *" = 4, " " = 1)) %>%
  row_spec(which(tab$Year == "Total"), bold=T) %>%
  add_footnote(label = "Most records in MPV and WA can be found in multiple sources, see other tabs for details on the overlap",
               notation = "symbol")

```

____

##### Primary source

When MPV includes an incident, that is our starting point for that case. Incidents from the other datasets are incorporated only when they fall outside of MPV's inclusion criteria.  In the table below, we show the *primary* source of cases by year:  all cases covered by MPV are listed in that column, the additional cases contributed from FE (2015-21), IN (2021+) and WA local (2022+) are shown in the other columns.

```{r tabAllCases}

tab <- homicides %>% 
  mutate(Source = factor(
    case_when(
      !is.na(mpvID) ~ "MPV",
      !is.na(inSource) ~ "IN",
      !is.na(feID) ~ "FE",
      !is.na(waID) ~ "WA",
      TRUE ~ "other"),
    levels = c("MPV", "FE", "IN", "WA"))
  ) %>%
  count(Source, Year = as.character(year)) %>% 
  pivot_wider(names_from=Source, values_from=n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.,0))) %>%
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  add_row(Year = "Total", summarise(., across(where(is.numeric), sum)))
  
tab %>%
  kableExtra::kbl(caption = "Primary data source by year") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_header_above(header = c(" " = 1, "Data Source *" = 4, " " = 1)) %>%
  row_spec(which(tab$Year == "Total"), bold=T) %>%
  add_footnote(label = "Most records in MPV and WA can be found in multiple sources, see other tabs for the overlap",
               notation = "symbol")
```


#### Source dataset descriptions {.tabset .tabset-pills}

##### Mapping Police Violence

*Data: 2013+*

[Mapping Police Violence (MPV)](https://mappingpoliceviolence.org/) aims to establish a comprehensive national open-source dataset of homicides by police, on and off duty.  It includes fatal shootings and fatalities caused by tasers, physical beatings, restraint holds, chemical spray, police dog attacks, less-lethal force and vehicular homicides.  Vehicular homicides are included in MPV *only when* the fatalities are directly caused by a police action: intentionally running someone over, and intentional collisions and vehicle disabling during a pursuit that leads to a fatal accident. As they note on their [methodology page](https://mappingpoliceviolence.org/methodology), these criteria result in the exclusion of most vehicle pursuit-related fatalities.  However they also note that they have begun to track these pursuits internally, so their coverage may change in the future. 


MPV has incorporated historical data from two legacy projects: 

* the [Fatal Encounters project](https://fatalencounters.org/) 2000-2021, and 

* the [Washington Post Police Fatal Force project]((https://www.washingtonpost.com/graphics/investigations/police-shootings-database/) 2015-2025.   

MPV merged and deduplicated the incidents from these two datasets, removed incidents that did not meet their inclusion criteria, added eligible incidents that were missed by both projects, kept the information from both datasets on the included incidents, and is now augmenting all incidents (back through 2013) with additional information, as well as updating with new incidents. 

MPV is a project of the [Campaign Zero](https://campaignzero.org/) organization.  It follows the principles of transparent reproducible research, with a systematic, multi-level review and coding process that is published on their [methodology page](https://mappingpoliceviolence.org/files/MappingPoliceViolence_Methodology.pdf) (recommended reading), and datasets provided in downloadable form to the public.

We download the MPV data from here:  https://mappingpoliceviolence.us/s/MPVDatasetDownload.xlsx.

____

##### IncarcerNation

*Data: 2021+*

IncarcerNation (IN) maintains a comprehensive, searchable website listing every fatality that has occurred during any police encounter in the United States, regardless of type, and without judgment. This includes deaths from direct police use of force [gunshots, tasers, K9, and other immediate lethal police action], as well as fatalities linked to police activities and interactions controlled by police that lead to death. These encompass high-speed pursuits (including deaths of drivers, passengers, bystanders, and pedestrians), collisions involving police vehicles, drownings during police chases, falls from heights during pursuit, asphyxiation during restraint, and suicides occurring in the context of police operations, often during SWAT incidents. ([taken from their "About Us" page](https://incarcernation.com/about-us/)).

IN's inclusion criteria are therefore much broader than MPV's, and they are the only actively updating project with coverage equivalent to the legacy Fatal Encounters project.  While most of their incidents overlap with MPV, they are the unique source for pursuit-related fatalities.

IN does not make their dataset downloadable for the public, so we manually review and transcribe information only for the cases they list for Washington state.  During transcription, duplicate cases, cases from Washington DC and cases that turned out to be non-police related homicides are excluded.  Many of the IN cases are also covered by MPV and the local WA project, but when IN is the primary source of information, we note that. 

____

##### Fatal Encounters

*Data: 2000-2021*

Fatal Encounters (FE) is a legacy project that sought to establish a comprehensive national open-source dataset of all deaths during encounters with police.  This project was active through December 31, 2021, includes records going back to 2000, and produced the largest compilation of persons killed during encounters with police that is publicly available: 31,641 fatalities at the time the project ended.  It covers deaths by all causes during a police encounter, including suicides, accidents, killings by off-duty police and vehicular pursuit related fatalities.  The vehicular pursuit coverage in FE is broader than MPV, as it includes all fatalities from accidents during a pursuit, so it captures bystanders killed by the fleeing subject.  

Legacy FE data are captured by the dataset used in this report in two ways:

* Cases included in the MPV dataset.

* Cases excluded by MPV are recovered and included by us.  They are used in this report, subject to the exclusions noted in a later section.

We originally downloaded the FE data from their website: https://docs.google.com/spreadsheets/d/1dKmaV_JiWcG8XBoRgP8b4e9Eopkpgt7FL7nyspvzAsE/edit?gid=0#gid=0 

We made many corrections to the FE data and harmonized the variables while FE was still updating.  The scripts for that can be found on our legacy GitHub repository: https://github.com/nextstepswa/fewapo.  We now just read in the cleaned, harmonized data (included in our current GitHub repository).  

____

##### Washington Post

*Data: 2015-2025*

The Washington Post Fatal Force (WaPo) dataset is a legacy project that sought to create a national database of fatal *shootings* by on-duty police.  It was actively updating from January 1, 2015 through December 31, 2024.  There were 10,429 fatalities recorded at the time the project ended.  

* WaPo historical data has been incorporated by the MPV project into their database, and is integrated into the WA local dataset via MPV.

The WaPo data can be downloaded from their GitHub repository here: https://github.com/washingtonpost/data-police-shootings.


____

For all of our data sources, it is worth noting that:

* None include deaths in custody after booking -- so all deaths in jails and prisons are excluded. 

* All sources may be missing incidents.

A comparison of the cases found in each dataset is presented in the 
[next section](#dataset-differences) of this report.

There has typically been a 1-2 week lag in MPV updates, but it can sometimes be longer.  The date shown in the "most recent data update" tab above is an indication of the current lag.  As of this report, the last data updates were:

* `r format(last_date_mpv, format="%B %d %Y")` for Mapping Police Violence
* `r format(last_date_in, format="%B %d %Y")` for IncarcerNation
* `r format(last_date_wa, format="%B %d %Y")` for the WA local data


This report is restricted to the cases that can be classified as homicides by police and vehicular homicides during police pursuits, it excludes cases where the cause of death during the police encounter is ruled a suicide (see the [next section](#dataset-differences) for details).

___
