### Discussion

Racial disparities in the risk of being killed by police are one of the most important factors driving the public demand for police accountability and reform.  For that reason it is important to understand how these numbers can, and cannot be used.

**TL;DR** While there are many uncertainties in the data that make it difficult/impossible to know the exact numbers, there are still some conclusions we can draw with confidence.


```{r include=F}

# Sources:
## https://www.ofm.wa.gov/washington-data-research/statewide-data/washington-trends/population-changes/population-race
## https://statisticalatlas.com/state/Washington/Race-and-Ethnicity

wa_ofm_pop_total <- readxl::read_xlsx(here::here("Data", "Raw", "ofm_pop_sade_state_2020_2024.xlsx"), sheet="Total")
wa_ofm_pop_hisp <- readxl::read_xlsx(here::here("Data", "Raw", "ofm_pop_sade_state_2020_2024.xlsx"), sheet="Hispanic")
wa_ofm_pop_nonhisp <- readxl::read_xlsx(here::here("Data", "Raw", "ofm_pop_sade_state_2020_2024.xlsx"), sheet="Non-Hispanic")

# 2020 Census WA populations by race
pop.tot <- wa_ofm_pop_total %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(Total) %>% as.numeric()
pop.w <- wa_ofm_pop_total %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(`White Total`) %>% as.numeric()
pop.b <- wa_ofm_pop_total %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(`Black Total`) %>% as.numeric()
pop.hisp <- wa_ofm_pop_hisp %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(`Hispanic Total`) %>% as.numeric()
pop.nhw <- wa_ofm_pop_nonhisp %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(`Non-Hispanic White Total`) %>% as.numeric()
pop.nhb <- wa_ofm_pop_nonhisp %>% filter(Year==2020 & `Age Group` == "Total") %>% 
  select(`Non-Hispanic Black Total`) %>% as.numeric()

# Proportions
prop.w.census <- pop.w/pop.tot
prop.b.census <- pop.b/pop.tot
prop.hisp.census <- pop.hisp/pop.tot
prop.nhw.census <- pop.nhw/pop.tot
prop.nhb.census <- pop.nhb/pop.tot

# Proportions for fatalities with race reported
hkr <- homicides %>% filter(race != "Unknown")
prop.wea.walocal <- prop.table(table(hkr$race))[["WEA"]]
prop.baa.walocal <- prop.table(table(hkr$race))[["BAA"]]
prop.hisp.walocal <- prop.table(table(hkr$race))[["HL"]]

# And proportions with unknown race
prop.wea.walocal.all <- prop.table(table(homicides$race))[["WEA"]]
prop.unk.walocal  <- prop.table(table(homicides$race))[["Unknown"]]

# Relative risks of being killed
rr.wea <- round(prop.wea.walocal/prop.nhw.census - 1, 2)
rr.baa <- round(prop.baa.walocal/prop.nhb.census - 1, 2)
rr.hisp <- round(prop.hisp.walocal/prop.hisp.census - 1, 2)
```

____

#### How is race identified in our data?

All crowd-sourced datasets rely primarily on media reports to find cases, and these media reports rarely include information on the race of the person killed.  Each of the projects that provide data for this report has made additional efforts made to identify race.

* MPV

The MPV project makes extensive additional efforts to identify the race of the person killed.  From 
[their data methodology page](https://mappingpoliceviolence.us/aboutthedata):

> Our data has been meticulously sourced from official police use of force data collection programs in states like California, Texas and Virginia, combined with nationwide data from The Gun Violence Archive and the Fatal Encounters database, two impartial crowdsourced databases. We've also done extensive original research to further improve the quality and completeness of the data; searching social media, obituaries, criminal records databases, police reports and other sources to identify the race of 90 percent of all victims in the database.


* Fatal Encounters 

The FE project used a statistical imputation model to predict race for the missing cases.  A brief description of the methodology is online [here](https://docs.google.com/spreadsheets/d/1dKmaV_JiWcG8XBoRgP8b4e9Eopkpgt7FL7nyspvzAsE/edit#gid=1359780841&range=A9).  They were able to impute just over half of the missing cases with reasonable confidence, and we use these imputations for the FE data in this report.


* Incarcernation

The IN project do not provide a description of the methods they use for coding race, but their records typically include information on the race of the person killed, as well as a photo.


* WA local

The local project conducts a search for coroner/ME reports, obituaries, multiple media reports and social media channels for information that can be used to assign the race of the person killed.


#### How often is race missing in our data?

We use a hierarchical assignment of race for the data in this report:  MPV data when available, IN data when available, WA local data for cases not included in MPV or IN, and Fatal Encounters (FE) with imputation for cases uniquely contributed by FE.

```{r racesource,  message=F, warning=F}
tab <- homicides %>%
  mutate(race = factor(race),
         Race = fct_relevel(race, "Unknown", after = Inf)) %>%
  group_by(Race, Source=source) %>%
  count()  %>%
  
  pivot_wider(names_from = Race, values_from = n) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>%

  add_row(Source = "Total", summarise(., across(where(is.numeric), sum))) %>%

  rowwise() %>%
  mutate(Total = sum(across(where(is.numeric))),
         "Unknown Pct" = round(100*Unknown/Total, 1)) %>%
  arrange(Total)


tab %>%
  kable(caption = "Breakdown by Source of Information") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(which(tab$source == "Total"), bold=T) %>%
  column_spec(which(names(tab)=="Total"), bold=T) %>%
  column_spec(which(names(tab)=="Unknown Pct"), italic=T) %>%
  add_footnote(label = "API = Asian and Pacific Islander; BAA = Black/African American; HL = Hispanic/Latino; NAA = Native American/Alaskan native; WEA = White/European American",
               notation = "symbol")
```

*Bottom line:* The unknown race cases make it impossible to say exactly how many people killed by police are in each racial group.

____

#### Reported vs. self-identified race

The race classification in our data represent the subject's race as reported by by some other person -- law enforcement, an ME/Coroner, a journalist, or a coder for one of the open-source projects that we draw from.  It may be that person's perception, or it may be informed by interactions with the subject's friends or family, we have no way of knowing. 

*Bottom line:*  The accuracy of the race classification in the data we have is unknown. 

____

#### We report the raw counts in this report, not per capita rates

Breaking the total count down by race, the largest single group of persons killed by police are identified as White/European-American: `r scales::percent(prop.table(table(homicides$race))[["WEA"]])`

So, does this mean that we can say: "There aren't any racial disparities in persons killed by police?".  
  
No. This raw count can not be used to assess racial disparities, because the word disparity implies the risk is "disproportionate" -- that is, higher (or lower) than proportional.  Proportional is a comparative term: it compares the proportion of fatalities by race, to the proportion of the population by race.  If the two proportions are the same, then we can say there are no disparities.  In this case, if `r scales::percent(prop.table(table(homicides$race))[["WEA"]])` of the WA state population is White, *only then* would we be able to say that their risk of being killed by police is proportional to their share of the population.

*Bottom line:* The raw counts can not be used to identify racial disparities.

____


#### Why don't we calculate standardized per capita rates?

We don't calculate standardized per capita rates because the race categories our dataset are not consistent with the race categories used by the US Census.  This is currently a limitation shared by all of the open-source datasets we rely on.  

Per capita rates are calculated by taking the ratio of the fatality count (in the numerator) to the population count (in the denominator).  This requires a consistent measure of race to use for the cases in the numerator and denominator - and we don't have that.  Census race categories are different for three reasons:

* The Census allows for multiple race coding:  About 5% of people in WA state report two or more races in the Census.  This type of multiple-race classification does not exist in the data from any of our source datasets, and it is almost never used in reports on persons killed by police, law enforcement statistics, or official medical examiner reports.  The likely reason for this is because the Census categories are self-reported -- by the persons themselves -- while race classification in the other settings is assigned by someone to another (deceased) person.

* The Census categories for single race are different: The standardized Census categories for race are American Indian/Alaska Native, Asian, Black, Native Hawaiian/Pacific Islander, White.  Slightly different versions are used in each of our source datasets.

* The Census separates the coding of race from ethnicity:  Hispanic/Latinx is an ethnicity that crosses several racial groups, primarily White, Black and Native American.  In the US Census data, ethnicity is measured separately from race, so one can observe both values for each person.  In all of our source datasets, "Hispanic" is coded as one of the racial classifications, rather than as a separate ethnicity classification, so the race of Hispanic persons can not be observed.

*For more information on how the Census codes race and ethnicity:*

[U.S. Decennial Census Measurement of Race and Ethnicity Across the Decades: 1790–2020](https://www.census.gov/library/visualizations/interactive/decennial-census-measurement-of-race-and-ethnicity-across-the-decades-1790-2020.html)

*Bottom line:* The mismatch between racial categories coded in our data, and the categories used in the US Census makes it impossible to calculate per capita rates accurately for all groups.

____


#### What can we say about racial disparities in the police use of deadly force in WA State?

We can say that, while we can not estimate the disparities with precision, or for all groups, the the data do suggest there are racial disparities in the police use of deadly force in WA State.  

The population of Washington State overwhelmingly identifies as White --  `r scales::percent(prop.w.census, accuracy=1)` in the last US Census -- while the fraction who identify as Black is `r scales::percent(prop.b.census, accuracy=1)`
(2020, 
[source: WA State OFM](https://ofm.wa.gov/data-research/population-demographics/estimates/age-sex-race-and-hispanic-origin/) for this and all other population estimates in this section).  The fraction (of all races) who report Hispanic ethnicity is 
`r scales::percent(prop.hisp.census, accuracy=1)`.  If we restrict the focus to non-Hispanics, these fractions drop to 
`r scales::percent(prop.nhw.census, accuracy=1)` and `r scales::percent(prop.nhb.census, accuracy=1)` for those who identify as White and Black respectively.

For these three groups -- non-Hispanic White, non-Hispanic Black and Hispanic -- our source data are coded in a way that we can compare our data to the Census populations, if we make three assumptions:

* First, that that cases classified as "White" or "Black" in the our datasets are not Hispanic -- because they would be classified as Hispanic otherwise -- and that cases classified as "Hispanic" in our dataset include all races.

* Second, that the coding of race and Hispanic origin in our dataset are accurate, and

* Third, that all race/Hispanic groups are equally likely to be coded as "unknown" race in our data. 

Under these assumptions (or, to the extent that they hold) we can compare the fraction of each group in the fatality data to their fraction in WA population.  The table below shows this, along with a summary based on the "risk ratio".

The "risk ratio" is the ratio of the fatality share to the population share.  When the two shares are the same, the risk ratio equals 1.  When the risk ratio is less than one, this means the share of fatalities is lower than the population share; the risks are disproportionately low.  When the risk ratio is greater than one it means the share of fatalities is larger than the population share; the risks are disproportionately high.  

This can be translated into a relative risk percentage: 100*(risk ratio-1), representing the percent lower (or higher) the risk ratio is than expected if the risk was proportional to the share of population. 

```{r race.tab, message=F, warning=F}

tab.all <- homicides %>%
  group_by(race) %>%
  summarize(all.cases = n()) %>%
  mutate(pct.all = round(100*all.cases/sum(all.cases), 1)) %>%
  filter(race %in% c("BAA", "HL", "WEA", "Unknown"))

tab.known <- homicides %>%
  filter(race != "Unknown") %>%
  group_by(race) %>%
  summarize(known.cases = n()) %>%
  mutate(pct.known = round(100*known.cases/sum(known.cases), 1)) %>%
  filter(race %in% c("BAA", "HL", "WEA")) %>%
  select(-known.cases)

tab.pop <- data.frame(race = c("BAA", "HL", "WEA"),
                      pct.pop = c(round(100*prop.nhb.census,1), round(100*prop.hisp.census, 1), 
                                  round(100*prop.nhw.census,1))
                      )

tab <- tab.all %>%
  left_join(tab.known) %>%
  left_join(tab.pop) %>%
  mutate(#rr.all = round((pct.all/pct.pop -1), 2),
         rr.known = round((pct.known/pct.pop - 1), 2)) %>%
  rename(Race=race, Fatalities=all.cases, "Percent (all cases)"=pct.all,
         "Percent (known race)"=pct.known, "Percent (WA pop)"=pct.pop, "Relative Risk" =rr.known) %>%
  mutate(Race = case_match(
    Race, "BAA" ~ "Black/African-American", 
    "HL" ~ "Hispanic/Latinx", 
    "WEA" ~ "White/European-American",
    .default = Race)
  )

tab %>% 
    kable(caption = "Police use of deadly force by race: Relative risks") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  add_footnote(label = "Relative risks are calculated based on percent of fatalities with known race",
               notation = "symbol")
```

In WA State:

  * For non-Hispanic Whites/European Americans this value is `r paste("100*(", scales::percent(prop.wea.walocal), "/", scales::percent(prop.nhw.census), "-1) =", rr.wea)` -- their risk of being killed by police is `r scales::percent(abs(rr.wea))` lower than expected. 
  
  * For non-Hispanic Blacks/African Americans this value is `r paste("100*(", scales::percent(prop.baa.walocal), "/", scales::percent(prop.nhb.census), "-1) = ", rr.baa)` -- their risk of being killed by police is `r scales::percent(rr.baa)` higher than expected.
  
  * For Hispanics this value is `r paste("100*(", scales::percent(prop.hisp.walocal), "/", scales::percent(prop.hisp.census), "-1) = ", rr.hisp)` -- their risk of being killed by police is `r scales::percent(rr.hisp)` higher than expected.

Note the direction of these disparities is robust to one of our assumptions:  that each group is equally likely to be reported as "race unknown".  Even if we assume that all of the "Unknown" race cases are non-Hispanic White, their share of incidents,
`r scales::percent(prop.wea.walocal.all + prop.unk.walocal, 1)`,
would still be below their share in the population.  And if all of the unknown cases are assumed to be either non-Hispanic Black or Hispanic, that would actually increase the relative risks for these groups. 

So, even though we can't be certain of the exact value of the risk ratio, we can say with some confidence that there *are* racial disparities in the risk of being killed by police, and that these disparities indicate that non-Hispanic Whites are *less* likely than other racial groups to be killed by law enforcement officers, while non-Hispanic Blacks and Hispanics are *more* likely to be killed. 

*Bottom line:* while the largest single group of persons killed by police in WA is identified as White, the risk ratio for this group, which adjusts for their share of the population, shows they are disproportionately **less** likely to be killed than those identified as Black or Hispanic.

___

#### Can we say anything about implicit bias?

Probably not.  To use these data as evidence for (or against) the effects of systemic implicit bias in the police use of deadly force, we would need to know the officer's *perception* of the subject's race, and control for other variables that may influence these patterns, like the perceived and actual weapon status of the person killed, the circumstances that precipitated the incidents, the presence or absence of other persons at risk in the incidents, as well as the demographic composition of the jurisdiction.

*Bottom line:* We can't use these data to answer the question of the officer's subjective intention or implicit bias.  But the data do support the argument that objective per capita disparities in risk by race exist.

____